---
title: "Harmful historical weather events in the US"
author: "Miguel Morales"
date: "December 27, 2015"
output: 
  html_document:
    keep_md: true
---

On this document, we will be briefly analyzing US weather data. We will be particularly
looking for information about harmful event since 1950 till recent years that have hit the country.

For this, we will be using [Government weather data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) freely available.

## Require Packages

There are a couple of packages that we will be using throughout this analysis. Here is the list of them:

```{r}
require(dplyr)
require(reshape2)
require(ggplot2)
require(gridExtra)
```

## Data Processing

This dataset presents several issues that we need to deal with it. Since we will working with 'dplyr', we will be converting the data frame into a table data frame object.

```{r}
df <- read.csv("repdata-data-StormData.csv.bz2")
data <- tbl_df(df)
```

First, lets take a look at the dataset dimensions and column names:

```{r}
dim(data)
names(data)
```

Let's have a little deeper look at the data:

```{r}
summary(data)
data
```

It seems that the most important columns are:

* EVTYPE
* FATALITIES
* INJURIES
* PROPDMG
* PROPDMGEXP
* CROPDMG
* CROPDMGEXP

Let's reduce the dataset to use only these columns:

```{r}
data <- select(data,
               EVTYPE,
               BGN_DATE,
               END_DATE,
               STATE,
               FATALITIES,
               INJURIES,
               PROPDMG,
               PROPDMGEXP,
               CROPDMG,
               CROPDMGEXP)
```

We need to make sure there are no missing values on the Event type column:

```{r}
any(is.na(data$EVTYPE))
```

No missing values on this columns. How about bad data?

```{r}
summary(data$EVTYPE)
(summarize(group_by(data, EVTYPE), n()) %>% arrange(desc(`n()`)))[1:20,]
```

There are several repeated Event Types, E.g.: TSTM WIND, THUNDERSTORM WIND, THUNDERSTORM WINDS. We will try to normalize these fields:

```{r}
data[data$EVTYPE == "TSTM WIND" | data$EVTYPE == "THUNDERSTORM WINDS", ]$EVTYPE = factor("THUNDERSTORM WIND")
data[data$EVTYPE == "MARINE TSTM WIND", ]$EVTYPE = factor("MARINE THUNDERSTORM WIND")
data[data$EVTYPE == "HURRICANE", ]$EVTYPE = factor("HURRICANE/TYPHOON")
data[data$EVTYPE == "RIVER FLOOD", ]$EVTYPE = factor("FLOOD")
```

Let's now fix the dates:
```{r}
data <- mutate(data, BGN_DATE = as.Date(BGN_DATE, format = "%m/%d/%Y"))
data <- mutate(data, END_DATE = as.Date(END_DATE, format = "%m/%d/%Y"))
```

It seems that PROPDMGEXP, and CROPDMGEXP have some issues:
```{r}
summary(data$PROPDMGEXP)
summary(data$CROPDMGEXP)
```

There should only be "K" for Thousands, "M" for Millions, "B" for Billions. Let's clean that up:
```{r}
data <- filter(data, PROPDMGEXP == "K" | PROPDMGEXP == "M" | PROPDMGEXP == "B" | PROPDMGEXP == "")
data <- filter(data, CROPDMGEXP == "K" | CROPDMGEXP == "M" | CROPDMGEXP == "B" | CROPDMGEXP == "")
data <- mutate(data, PROPDMGEXP = factor(PROPDMGEXP, levels = c("", "K", "M", "B")))
data <- mutate(data, CROPDMGEXP = factor(CROPDMGEXP, levels = c("", "K", "M", "B")))
data[data$PROPDMGEXP == "K",] <- filter(data, PROPDMGEXP == "K") %>% mutate(PROPDMG = PROPDMG * 1000)
data[data$PROPDMGEXP == "M",] <- filter(data, PROPDMGEXP == "M") %>% mutate(PROPDMG = PROPDMG * 1000000)
data[data$PROPDMGEXP == "B",] <- filter(data, PROPDMGEXP == "B") %>% mutate(PROPDMG = PROPDMG * 1000000000)
data[data$CROPDMGEXP == "K",] <- filter(data, CROPDMGEXP == "K") %>% mutate(CROPDMG = CROPDMG * 1000)
data[data$CROPDMGEXP == "M",]  <- filter(data, CROPDMGEXP == "M") %>% mutate(CROPDMG = CROPDMG * 1000000)
data[data$CROPDMGEXP == "B",]  <- filter(data, CROPDMGEXP == "B") %>% mutate(CROPDMG = CROPDMG * 1000000000)
tidy_data <- select(data,
               -PROPDMGEXP,
               -CROPDMGEXP)
```

I think this is a much cleaner data set, let's find answers to our questions.
```{r}
summary(tidy_data)
tidy_data
```



## Results
There should be a section titled Results in which your results are presented.

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Let's explore the data into this question:
```{r}
health <- summarize(group_by(tidy_data, EVTYPE), sum(FATALITIES), sum(INJURIES))
names(health) <- c("type", "fatalities", "injuries")
top_health <- arrange(health, desc(fatalities), desc(injuries))[1:10,]
top_health
```

Let's plot this:
```{r fig.width=10, fig.height=6,}
fatal_plot <- ggplot(top_health, aes(x = type, y = fatalities)) + geom_bar(stat="identity") +
xlab("Event Type") + ylab("Total Fatalities") + ggtitle("Total Fatalities per Event Type") +
theme(axis.text.x = element_text(angle = 90, hjust = 1))

injure_plot <- ggplot(top_health, aes(x = type, y = injuries)) + geom_bar(stat="identity") +
xlab("Event Type") + ylab("Total Injuries") + ggtitle("Total Injuries per Event Type") +
theme(axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(fatal_plot, injure_plot, ncol=2)
```

It seems that Tornados are the biggest cause of personal health issues in the US.

2. Across the United States, which types of events have the greatest economic consequences?

Let's explore the data into this question:
```{r}
damage <- summarize(group_by(tidy_data, EVTYPE), sum(PROPDMG), sum(CROPDMG))
names(damage) <- c("type", "property", "crop")
damage <- mutate(damage, total = property + crop)
top_damage <- arrange(damage, desc(total), desc(property), desc(crop))[1:10,]
top_damage <- select(top_damage, type, property, crop)
top_damage <- mutate(top_damage, property = property / 1000000)
top_damage <- mutate(top_damage, crop = crop / 1000000)
top_damage <- melt(top_damage, id.var="type")
top_damage
```

And the plot that would give us the answer.
```{r fig.width=10, fig.height=6,}
ggplot(top_damage, aes(x = type, y = value, fill = variable)) + 
geom_bar(stat="identity") +
scale_fill_discrete(name="Damage Type") +
xlab("Event Type") + ylab("Total Damage (in Millions of Dollars)") + 
ggtitle("Total Damage per Event Type") +
theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

It seems that Flood is the biggest cause of money damage on a weather event.

